Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-02-07 paper/paper.tex
resolved bad wording and added a cifar10 figure

The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks.
%
\todo{reword}
We observe dual effect of ``fine-tune'' stage: it improves performance of highly compressed
models since they are more regularized (by poking holes) and thus serves as further fitting.
At the same time, undercompressed models, behave as if they were over parameterized
and the fine tune stage overfits.

% on compresssion
Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the $q_\theta$ factorizes into univariate distributions, a $\cplx$ or $\real$ parameter is
considered non-zero if $\log \alpha \leq \tau$ for $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$, or zero otherwise. Figure~\ref{fig:hist__and__threshold__tradeoff} (top) displays
the general shape of the distribution of $\log \alpha$ on the Fashion-MNIST dataset in
sec.~\ref{sub:mnist_like_datasets}. There always is a single mode within the moderately
negative range, and one or two modes to the right of zero, and the trough between the
peaks of the negative mode and the first positive one is around zero. The relative mass
of each mode changes with $C$ and the higher the coefficient is, the more mass is shifted
to the right.
%
The bottom graph on figure~\ref{fig:hist__and__threshold__tradeoff} is constructed
for $C \in \{\tfrac1{20}, \frac1{200}\}$ before running the ``fine-tune'' stage. From
\eqref{eq:elbo_with_coef}, the relative positions of the compression curves on this
plot and the results of other experiments, it can be concluded that $C$ has much more
substantial impact on the sparsity and performance, than the choice of $\tau$. Furthermore,
for $\tau > 0$ the performance quickly saturates and while compression continues to
drop.

\begin{figure}[ht]
  \centering
  \begin{subfigure}[b]{0.95\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\linewidth]{../assets/log-alpha__sum__fashion-mnist__20200131-205926.pdf}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{0.95\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\linewidth]{../assets/figure__musicnet__threshold.pdf}
  \end{subfigure}
  \caption{%
    (\textit{top}) Smoothed density of $\log\alpha$ for some convolutional and dense layer
    from a model in sec.~\ref{sub:mnist_like_datasets};
    (\textit{bottom}) Test split performance and compression rate for $\cplx$-valued net
    by \citet{trabelsi_deep_2017} on MusicNet.
  }
  \label{fig:hist__and__threshold__tradeoff}
\end{figure}

% on chosing the threshold
The threshold $\tau$ is picked so, that the remaining \textit{``non-zero''} parameters
are within $\delta$ relative tolerance of their mean with high probability under the
approximate posterior $q_\theta(W)$. This ensures that stochastic parameters are sufficiently
concentrated around $\mu$, which is used as a stand-in value for $W$, when the network
is used in inference mode. Since $W$ is $\real$- or a proper $\cplx$-Gaussian, the random
variable $
  \tfrac{k \lvert W - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    % = \Pr\biggl(
    %   \frac{k \lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
    %     \leq \frac{\delta^2 k}{\alpha}
    % \biggr)
    = \chi^2_k\biggl(\frac{\delta^2 k}{\alpha}\biggr)
  \,.
\end{equation}
For a generous $\delta = 50\%$ tolerance all values of $\log \alpha$ below $-2.5$ yield
at least $90\%$ chance of a non-zero $\real$/$\cplx$ parameter. We pick $\tau = -\tfrac12$
to exclude noisy parameters and encourage higher sparsity, at the same time staying aware
that $q_\theta$ is merely an approximation. In comparison, $\tau = 3$ is commonly used as
the threshold \citep{molchanov_variational_2017,kingma_variational_2015}.

@@

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is computed based on the number of floating point values needed
to specify the model and equals $
  \frac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros after sparsification, $n_\mathtt{par}$
is the total number of floats. Thus each parameter is counted once in a $\real$-valued
network, but twice in a $\cplx$-network, being a pair of real-imaginary floats. Each model
has incompressible parameters, such as layers' biases, shift and scaling from $\real$-and
$\cplx$-batch normalization layers, if present.

@@

% a couple of plots from MNIST and Fashion-MNIST
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__fashionmnist__fft__-0.5.pdf}
    \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__kmnist__fft__-0.5.pdf}
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/VD__kmnist__fft__-0.5.pdf}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__mnist__fft__-0.5.pdf}
    \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__emnist_letters__fft__-0.5.pdf}
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/ARD__kmnist__fft__-0.5.pdf}
  \end{subfigure} \\%
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__fashionmnist__raw__-0.5.pdf}
    \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__kmnist__raw__-0.5.pdf}
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/VD__kmnist__raw__-0.5.pdf}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__mnist__raw__-0.5.pdf}
    \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__emnist_letters__raw__-0.5.pdf}
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/VD__emnist_letters__raw__-0.5.pdf}
  \end{subfigure}
  \caption{%
    The compression-accuracy trade-off after ``fine-tune'' on MNIST of \textit{TwoLayerDenseModel}
    (top) and \textit{SimpleConvModel} (bottom) for methods in sections~\ref{ssub:variational_dropout}
    and~\ref{ssub:empirical_bayes}.
  }
  \label{fig:mnist-like__trade-off}
\end{figure}

@@

% mention the same experimental setup as in MNIST
Experiments on CIFAR10 dataset are set up similarly to MNIST-like experiment, except
that we study VGG16 model \citep{simonyan_very_2015} and its $\cplx$-counterpart. Each
network is trained stagewise for $20$, $40$, and $20$ epochs on the full training split
of $50k$ colour $32\times32$ images. We use only raw colour features (\texttt{raw}), since
the results of the experiments on MNIST left us sceptical about the utility of the Fourier
features for image recognition. We use fast learning rate scheduler from sec.~\ref{sub:mnist_like_datasets}.

% mention that fourier-features perfomed worse than raw image data embeded in to $\cplx$

% describe the VGG16 model and its cplx-valued modification
The $\cplx$-VGG16 model is architecturally identical to VGG16, except for all layers
replaced by their $\cplx$-valued versions. To maintain comparability, we halve the
filter sizes of $\cplx$ model \citep{monning_evaluation_2018}.

% compare compressed cplx-vgg against real-vgg
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=\linewidth]{../assets/figure__cifar10__trade-off/musicnetram__fft__-0.5.pdf}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=\linewidth]{../assets/figure__cifar10__threshold/cifar10__threshold.pdf}
  \end{subfigure}
  \caption{%
    Pre ``fine-tune'' performance of the best models from figure \ref{fig:musicnet_trade-off}
    compressed with methods of sec.~\ref{sec:dropout_for_c_valued_layers}:
    (\ref{ssub:variational_dropout} -- \textit{top}, \ref{ssub:empirical_bayes} --  \textit{bottom}).
  }
  \label{fig:figure__cifar10__trade-off}
\end{figure}

@@

@@

% describe musicnet as in Trabelsi et al. (2017) also get some hints from thickstun_learning_2017
MusicNet \citep{thickstun_learning_2017} is an annotated audio dataset, consisting of $330$
compositions from the public domain. Folowing \citet{trabelsi_deep_2017}, we use this
dataset for automatic music transcription task and study the $1d$ VGG-like $\cplx$-convent
proposed therein. Similarly to this study, we downsample the sound from 44.1kHz to 11kHz,
retaining only $84$ out of $128$ labels, and hold-out the same validation and test splits.
During each training epoch we sample $1000$ mini-batches of the remaining $321$ compositions
thus: a random full window of $4096$ samples is picked within each waveform and transformed
from $\real$ to $\cplx$ with $1d$ FFT. Annotation labels are taken at the mid-point of the
window.

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\linewidth]{../assets/figure__musicnet__trade-off/musicnetram__fft__-0.5.pdf}
  \end{subfigure}
  \caption{%
    \textbf{\color{blue} VD} and \textbf{\color{orange} ARD}
    denote
    methods \ref{ssub:variational_dropout}) and \ref{ssub:empirical_bayes}, respectively.
    The \textbf{\color{pink} version} of the network with kernel $3$ in every convolution compressed with VD
    is (\ref{ssub:variational_dropout}).
  }
  \label{fig:musicnet__trade-off}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\linewidth]{../notebooks/assets/grad_log.pdf}
  \caption{$\tfrac{d f(\alpha)}{d \log{\alpha}}$ of the fit proposed in
  \citet{molchanov_variational_2017}, MC estimate of \eqref{eq:improper-kl-div-real},
  and the exact derivative using \eqref{eq:real-kl-div-deriv-log}.}
  \label{fig:molchanov-derivative-replica}
\end{figure}

>>>>

The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks. 
%
The benefit of the ``fine-tune'' stage appears to be strongly influenced by the compression
rate. In this experiment and more prominently in MusicNet (sec.~\ref{sub:musicnet}), we
can observe that the more ``holes'' Bayesian dropout poked in the model, the more regularized
it becomes, the less capacity it maintains for overfitting and the more ``fine-tune'' serves
as extended training.

% on compresssion
Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the $q_\theta$ factorizes into univariate distributions, a $\cplx$ or $\real$ parameter is
considered non-zero if $\log \alpha \leq \tau$ for $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$, or zero otherwise. Figure~\ref{fig:hist__and__threshold__tradeoff} (top) displays the
general shape of the distribution of $\log \alpha$ on the Fashion-MNIST dataset in
sec.~\ref{sub:mnist_like_datasets}. There always appears to be a mode in the moderately
negative range, a valley through zero between the peaks of the negative mode and the next
positive one. The relative mass of each mode changes with $C$ and the higher the coefficient
is, the more mass is shifted to the right.
%
The bottom graph on figure~\ref{fig:hist__and__threshold__tradeoff} is constructed
for $C \in \{\tfrac1{20}, \frac1{200}\}$ before running the ``fine-tune'' stage. From
\eqref{eq:elbo_with_coef}, the relative positions of the compression curves on this
plot and the results of experiments below, it can be concluded that $C$ has much more
substantial impact on the sparsity and performance, than the choice of $\tau$. Furthermore,
for $\tau > 0$ the performance quickly saturates while compression creeps downwards.

\begin{figure}[ht]
  \centering
  \begin{subfigure}[b]{0.95\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\linewidth]{../assets/log-alpha__sum__fashion-mnist__20200131-205926.pdf}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{0.95\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\linewidth]{../assets/figure__musicnet__threshold.pdf}
  \end{subfigure}
  \caption{%
    (\textit{top}) Smoothed density of $\log\alpha$ for some convolutional and dense layer
    from a model in sec.~\ref{sub:mnist_like_datasets};
    (\textit{bottom}) Test split performance and compression rate for a $\cplx$-valued
    net on MusicNet.
  }
  \label{fig:hist__and__threshold__tradeoff}
\end{figure}

% on chosing the threshold
The threshold $\tau$ is picked so, that the remaining \textit{``non-zero''} parameters
are within $\delta$ relative tolerance of their mean with high probability under the
approximate posterior $q_\theta(W)$. This ensures that stochastic parameters are sufficiently
concentrated around $\mu$, which are used as stand-in values for network weights $W$,
when it runs in inference mode. Since $W$ is $\real$- or a proper $\cplx$-Gaussian, the
random variable $
  \tfrac{k \lvert W - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    % = \Pr\biggl(
    %   \frac{k \lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
    %     \leq \frac{\delta^2 k}{\alpha}
    % \biggr)
    = \chi^2_k\biggl(\frac{\delta^2 k}{\alpha}\biggr)
  \,.
\end{equation}
For a generous $\delta = 50\%$ tolerance all values of $\log \alpha$ below $-2.5$ yield
at least $90\%$ chance of a non-zero $\real$/$\cplx$ parameter. We pick $\tau = -\tfrac12$
to exclude noisy parameters and encourage higher sparsity, at the same time being aware
that $q_\theta$ is merely an approximation. In comparison, $\tau = 3$ is commonly used as
the threshold \citep{molchanov_variational_2017,kingma_variational_2015}.

@@

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is computed based on the number of floating point values needed to
specify the model and equals $
  \tfrac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros after sparsification, $n_\mathtt{par}$
is the total number of floats. Thus each parameter is counted once in a $\real$-valued
network, but twice in a $\cplx$-network, being a pair of real-imaginary floats. Each model
has incompressible parameters, such as layers' biases, shift and scaling from $\real$-and
$\cplx$-batch normalization layers, if present.

@@

% a couple of plots from MNIST and Fashion-MNIST
\begin{figure}[!h]
  \centering
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__fashionmnist__fft__-0.5.pdf}
    \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__kmnist__fft__-0.5.pdf}
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/VD__kmnist__fft__-0.5.pdf}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__mnist__fft__-0.5.pdf}
    \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__emnist_letters__fft__-0.5.pdf}
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/ARD__kmnist__fft__-0.5.pdf}
  \end{subfigure} \\%
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__fashionmnist__raw__-0.5.pdf}
    \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__kmnist__raw__-0.5.pdf}
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/VD__kmnist__raw__-0.5.pdf}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__mnist__raw__-0.5.pdf}
    \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/legacy__VD__emnist_letters__raw__-0.5.pdf}
    % \includegraphics[width=\linewidth]{../assets/figure__mnist-like__trade-off/VD__emnist_letters__raw__-0.5.pdf}
  \end{subfigure}
  \caption{%
    The compression-accuracy trade-off after ``fine-tune'' on MNIST of \textit{TwoLayerDenseModel}
    (top) and \textit{SimpleConvModel} (bottom) for methods in sections~\ref{ssub:variational_dropout}
    and~\ref{ssub:empirical_bayes}.
  }
  \label{fig:mnist-like__trade-off}
\end{figure}

@@

% mention the same experimental setup as in MNIST
Experiments on CIFAR10 dataset are set up similarly to MNIST-like experiment, except
that we study VGG16 model \citep{simonyan_very_2015} and its $\cplx$-counterpart. Each
network is trained stagewise for $20$, $40$, and $20$ epochs on the full training split
of $50k$ colour $32\times32$ images. We use only raw colour features (\texttt{raw}), since
the results of the experiments on MNIST left us sceptical about the utility of the Fourier
features for image recognition. We use fast learning rate scheduler from sec.~\ref{sub:mnist_like_datasets}
and apply random crop and horizontal flips as augmentation.

% describe the VGG16 model and its cplx-valued modification
The $\cplx$-VGG16 model is architecturally identical to VGG16, except for all layers are
replaced by their $\cplx$-valued versions. To maintain comparability, we halve the filter
sizes of $\cplx$ model \citep{monning_evaluation_2018}.

@@

% compare compressed cplx-vgg against real-vgg
\begin{figure}[!h]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\linewidth]{../assets/figure__cifar__trade-off/augmentedcifar10__raw__-0.5.pdf}
  \end{subfigure}
  \caption{%
    Compression-accuracy frontier for $\real$ and $\tfrac12 \cplx$ versions of VGG16 on CIFAR10.
  }
  \label{fig:figure__cifar10__trade-off}
\end{figure}

\begin{figure}[!h]
  \centering
  \begin{subfigure}[b]{1.\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\linewidth]{../assets/figure__musicnet__trade-off/musicnetram__fft__-0.5.pdf}
  \end{subfigure}
  \caption{%
    Performance-compression trade-off using \textbf{\color{blue} VD} and \textbf{\color{orange} ARD}.
    The \textbf{\color{pink} version} of the network with kernel $3$ in the first convolution is
    compressed with VD only.
  }
  \label{fig:musicnet__trade-off}
\end{figure}

@@

% describe musicnet as in Trabelsi et al. (2017) also get some hints from thickstun_learning_2017
MusicNet \citep{thickstun_learning_2017} is an annotated audio dataset, consisting of $330$
compositions from the public domain. Folowing \citet{trabelsi_deep_2017}, we use this
dataset for automatic music transcription task and study the $1d$ VGG-like $\cplx$-convent
proposed therein. Similarly to this study, we downsample the audio from 44.1kHz to 11kHz,
retaining only $84$ out of $128$ labels, and hold-out the same validation and test splits.
During each training epoch we sample $1000$ mini-batches of the remaining $321$ compositions
thus: a random full window of $4096$ samples is picked within each waveform and transformed
from $\real$ to $\cplx$ with $1d$ FFT. Annotation labels are taken at the mid-point of the
window.

@@

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=\linewidth]{../notebooks/assets/grad_log.pdf}
  \caption{$\tfrac{d f(\alpha)}{d \log{\alpha}}$ of the fit proposed in
  \citet{molchanov_variational_2017}, MC estimate of \eqref{eq:improper-kl-div-real},
  and the exact derivative using \eqref{eq:real-kl-div-deriv-log}.}
  \label{fig:molchanov-derivative-replica}
\end{figure}
